---
title: "Proteomics analysis"
author: chripla + grst
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
params:
  maxcores: 6
  inputdir: "../../data/proteomics"
  inputdir_clinical: "../../data"
  resultdir:  "../../results/proteomics"
  paper_figures: "../../results/Figures"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(conflicted)
library(dplyr)
library(tibble)
library(ggbiplot)
library(factoextra)
library(tidyr)
library(ggrepel)
library(clusterProfiler)
library(org.Hs.eg.db)
library(cowplot)
library(ComplexHeatmap)
library(RColorBrewer)
library(readxl)
library(biomaRt)
library(reshape2)
library(tidyverse)
library(parallel)
library(msigdbr)
library(GSVA)
conflict_prefer("crossing", "tidyr")
conflict_prefer("get_legend", "ggpubr")
conflict_prefer("filter",  "dplyr")
conflict_prefer("select",  "dplyr")
conflict_prefer("mutate",  "dplyr")
conflict_prefer("rename", "dplyr")
source("../../lib/global_params.R") # import color-codes
```


```{r, include=FALSE}
## Function for excluding non distinct proteins:
exclude_proteins <- function(prot, genes = FALSE) {
  y <- grep("[^0-9]", (substring(prot$Protein, 1, 1)))
  prot <- prot[y, ]
  
  if (genes == TRUE) {
    # exclude non distinct genes:
    prot = prot[!is.na(prot$`Gene name`),]
    prot = prot[prot$`Gene name` != '', ]
    prot <- prot[!duplicated(prot$`Gene name`),] |>
      column_to_rownames("Gene name")
  } else {
    rownames(prot) <- prot$Protein
  }
  prot <- prot[, startsWith(names(prot), "CRC")]
  return(prot)
}
```


### Load files:
1. analysis output for all organoids (with and without imputation) including intensity values for all replicates
2. clinical sample annotation

```{r}
##1
file.prot.avg.wImp <- file.path(params$inputdir, "protein_level_baseline_wImp_averaged.txt")
file.prot.avg <- file.path(params$inputdir, "protein_level_baseline_woImp_averaged.txt")
file.prot.wImp <- file.path(params$inputdir, "protein_level_baseline_wImp.txt")

prot.avg <- read_tsv(file.prot.avg)
prot.avg.wImp <- read_tsv(file.prot.avg.wImp)
prot.wImp <- read_tsv(file.prot.wImp)

##2
clinical <- read_tsv(file.path(params$inputdir_clinical, "clinical_data_organoids.tsv"))
```


Reshape data and exclude non-distinct proteins:
```{r}
# For the file with replicates, add gene names from annotation table:
prot_to_genename <- read_tsv("../../tables/prot_to_gene_name.tsv")
prot.wImp <- prot.wImp |>
  left_join(prot_to_genename) |>
  select(-c('nFragment', 'nPeptide'))

# exclude non distinct proteins:
prot.avg <- exclude_proteins(prot.avg, genes=TRUE)
prot.avg.wImp <- exclude_proteins(prot.avg.wImp, genes=TRUE)
prot.wImp <- exclude_proteins(prot.wImp, genes=TRUE)

# ordered list according to mutations:
samples_ordered <- c("CRC22", "CRC19", "CRC13", "CRC01", "CRC15", "CRC04", "CRC03", "CRC02", "CRC12",  "CRC26LM", "CRC17", "CRC20", "CRC26", "CRC11", "CRC18", "CRC21")
prot.avg <- prot.avg[, match(samples_ordered, names(prot.avg))] 
prot.avg.wImp <- prot.avg.wImp[, match(samples_ordered, names(prot.avg.wImp))]

# set missing values to NA in the non-imputed data frame:
prot.avg[prot.avg==0]<-NA

# transform to log2-scale:
prot.avg <- log2(prot.avg +1)
prot.avg.wImp <- log2(prot.avg.wImp+1)
prot.wImp <- log2(prot.wImp +1)

#rename CRC03_L to CRC03:
names(prot.wImp) <- gsub("L_", "", names(prot.wImp))
```

### PCA from averaged replicates:

```{r, fig.width=6, fig.height=4.5}
prot.pca <- na.omit(prot.avg.wImp)
prot.pca <- as.data.frame(t(prot.pca))

prot.pca <- prcomp(prot.pca)
prot.pca.df <- prot.pca$x |> as_tibble(rownames="sample") |> separate(sample, "organoid", fill="right", extra="merge", sep="_", remove=FALSE)
prot.pca.df$organoid <- sub("_.*", "", prot.pca.df$organoid)

ggbiplot(prot.pca, var.axes = FALSE,  groups=prot.pca.df$organoid) +
           geom_text_repel(label = prot.pca.df$sample) + #add labels non-overlapping (ggrepel-package)
           theme_bw()+
           theme(legend.position = "none")+
           scale_color_manual(values=colors$organoid) + 
           ggtitle("PCA - Proteomics (with imputation)")

```

Cutoff for GO analysis:
```{r, fig.width=8, fig.height=6}
plot(sort(prot.pca$rotation))

plot(sort(prot.pca$rotation[,1]))

# take +/- 0.025 as cut-off
prot.pc1 <- prot.pca$rotation[abs(prot.pca$rotation[,1]) >= 0.025, ]
dim(prot.pc1)
```
GO-analysis - molecular function:
```{r}
genes.pc1 <- rownames(prot.pc1)

go.pc1 = enrichGO(gene=genes.pc1, OrgDb=org.Hs.eg.db, keyType="SYMBOL", ont = "MF", pAdjustMethod = "BH", pvalueCutoff = 0.01, qvalueCutoff = 0.05)
go.pc1@result <- go.pc1@result |> filter(p.adjust < 1) |> bind_rows() |> mutate(abs_log_p = -log10(p.adjust))

pc1_top_go_mf = go.pc1  |> mutate(rk = rank(pvalue, ties.method="min")) |> filter(rk <= 10)
pc1_top_go_mf@result
```


### Heatmap of all proteins with GO analysis

The following code snippets produce a clustered heatmap of all proteins. 
 * all-zero proteins are excluded
 * The protein levels are z-scaled and clustered hierarchically by using Pearson correlation as distance and complete linkage. 
 * 8 Major clusters are defined and annotated using GO enrichment analysis. 

```{r, fig.width=4, fig.height=7}
tpm_mat2 = prot.avg.wImp
colnames(tpm_mat2) <- sub("_.*", "", colnames(tpm_mat2))

tpm_mat_scaled =t(scale(t(tpm_mat2[!apply(tpm_mat2, 1, function(row) {all(is.na(row))}),])))

dist_obj = factoextra::get_dist(tpm_mat_scaled, "pearson")
clust_obj = fastcluster::hclust(dist_obj)
```

Run GO analysis: 
```{r}
N_CLUSTERS = 8
clusters = cutree(clust_obj, k=N_CLUSTERS)
universe = rownames(tpm_mat_scaled)

# dummy see https://github.com/YuLab-SMU/clusterProfiler/issues/207
.x = enrichGO(c("CD8A"), OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP", universe = universe)

# molecular function
ego_mf = mclapply(unique(clusters), function(clus) {
  genes = names(which(clusters == clus))
  ego = enrichGO(gene=genes, OrgDb=org.Hs.eg.db, keyType="SYMBOL", ont = "MF", pAdjustMethod = "BH", pvalueCutoff = 0.01, qvalueCutoff = 0.05, universe=universe)
  ego@result$cluster =clus
  ego
}, mc.cores=32)

# biological process
ego_bp = mclapply(unique(clusters), function(clus) {
  genes = names(which(clusters == clus))
  ego = enrichGO(gene=genes, OrgDb=org.Hs.eg.db, keyType="SYMBOL", ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.01, qvalueCutoff = 0.05, universe=universe)
  ego@result$cluster =clus
  ego
}, mc.cores=32)
```
Combine into dataframe: 
```{r}
# for some categories there are no enriched GO terms. Therefore the tryCatch. 
go_mf = lapply(ego_mf, function(ego) {
  tryCatch({
    ego@result %>% filter(p.adjust < 1)
  }, error = function(e) {tibble()})
}) %>% bind_rows() %>% mutate(abs_log_p = -log10(p.adjust))


go_bp = lapply(ego_bp, function(ego) {
  tryCatch({
    ego@result %>% filter(p.adjust < 1)
  }, error = function(e) {tibble()})
}) %>% bind_rows() %>% mutate(abs_log_p = -log10(p.adjust))
```

```{r}
# Limit to 5 terms per cluster
top_go_mf = go_mf |> group_by(cluster) |> mutate(rk = rank(pvalue, ties.method="min")) |> filter(rk <= 5) |> ungroup()
top_go_bp = go_bp |> group_by(cluster) |> mutate(rk = rank(pvalue, ties.method="min")) |> filter(rk <= 5) |> ungroup()
```

Visualize GO terms: 
```{r, fig.width=15, fig.height=15}
p_mf = top_go_mf |> select(Description, abs_log_p, cluster) |> 
  right_join(crossing(cluster=unique(clusters), Description=top_go_mf$Description)) |>
  mutate(cluster = as.character(cluster)) |>
  ggplot(aes(y=Description, x=cluster, fill=abs_log_p)) + geom_tile(col='grey') + theme_cowplot() +
  ylab("Molecular function") + scale_fill_distiller("-log10(FDR)", palette="YlGnBu", direction=1, na.value="white", limits=c(0, 15)) +
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_bp = top_go_bp %>% select(Description, abs_log_p, cluster) |> 
  right_join(crossing(cluster=unique(clusters), Description=top_go_bp$Description)) |>
  mutate(cluster = as.character(cluster)) |> 
  ggplot(aes(y=Description, x=cluster, fill=abs_log_p)) + geom_tile(col='grey') + theme_cowplot() + 
  ylab("Biological process") + scale_fill_distiller("-log10(FDR)", palette="YlGnBu", direction=1, na.value="white", limits=c(0, 15)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

legend = get_legend(p_mf)
p_mf = p_mf + theme(legend.position = "none")
p_bp = p_bp + theme(legend.position = "none")
plot_grid(
  plot_grid(p_mf, p_bp, ncol=1, align="v", rel_heights = c(.3, .45)),
  legend,
  nrow=1, 
  rel_widths = c(.8, .2)
)
```

Produce heatmap: 
```{r, fig.height=4.5, fig.width=5.5}
cluster_annotations = list(
  "1"="protein localization\nto plasma membrane",
  "2"="protein localization\nto chromosome",
  "3"="metabolic processes",
  "4"="translational activity",
  "5"="RNA processing",
  "6"="extracellular\nmatrix organization",
  "7"="",
  "8"="exocytosis"
)
annotation_df = data.frame(cluster = as.character(clusters))

ha = rowAnnotation(clusters=anno_empty(border=FALSE, width=unit(50, 'mm')))

Heatmap(tpm_mat_scaled, 
        left_annotation=ha, 
        cluster_rows=clust_obj,
        cluster_columns =TRUE,
        show_row_names = FALSE,
        show_row_dend = FALSE,
        show_column_dend = FALSE,
        clustering_distance_columns = "pearson",
        row_title = NULL,
        split = N_CLUSTERS,
        name='z-score',
        use_raster = TRUE)

palette(brewer.pal(N_CLUSTERS, "Set2"))
lapply(names(cluster_annotations), function(c) {
  decorate_annotation("clusters", slice = c, {
    grid.rect(x=1, width=unit(2, "mm"), just="right", gp = gpar(fill = c, col=NA))
    grid.text(cluster_annotations[[c]], just="right", x=unit(45, "mm"), gp=gpar(fontsize=12, lineheight=1))
  })
})
```

Export heatmap as Supp. Figure 1E:
```{r}
pdf(file.path(params$paper_figures, "SuppFigure1E.pdf"), width = 8, height = 9, pointsize = 30)
Heatmap(tpm_mat_scaled, 
        left_annotation=ha, 
        cluster_rows=clust_obj,
        cluster_columns =TRUE,
        show_row_names = FALSE,
        show_row_dend = FALSE,
        show_column_dend = FALSE,
        clustering_distance_columns = "pearson",
        row_title = NULL,
        split = N_CLUSTERS,
        name='z-score',
        use_raster = TRUE)

palette(brewer.pal(N_CLUSTERS, "Set2"))
lapply(names(cluster_annotations), function(c) {
  decorate_annotation("clusters", slice = c, {
    grid.rect(x=1, width=unit(2, "mm"), just="right", gp = gpar(fill = c, col=NA))
    grid.text(cluster_annotations[[c]], just="right", x=unit(45, "mm"), gp=gpar(fontsize=12, lineheight=1))
  })
})
dev.off()
```

### GSVA with HALLMARK gene sets

Get HALLMARK gene sets from migSigDB (version: 7.5.1):
```{r}
h <- msigdbr(species = "Homo sapiens", category = "H") |> dplyr::select(gene_symbol, gs_name) |> unstack()
``` 

Calculate GSVA for all samples and take mean per organoid:
```{r HALLMARK heatmap}
gsva_results <- gsva(data.matrix(prot.wImp), h, kcdf = "Gaussian", parallel.sz = params$maxcores)

heatdata = gsva_results |>
  as.data.frame() |>
  rownames_to_column("hallmark") |>
  pivot_longer(cols=starts_with("CRC"), names_to = "sample") |> 
  separate("sample", into=c("organoid", "replicate"), extra = "merge") |>
  group_by(hallmark, organoid) |>
  dplyr::summarise(value=mean(value)) |>
  pivot_wider(id_cols="hallmark", names_from="organoid") |>
  column_to_rownames('hallmark')

heatdata <- t(scale(t(heatdata)))

top_df <- data.frame(row.names = colnames(heatdata))
top_df$`MSI status` <- clinical$`MSI / MSS (IHC)`[match(rownames(top_df), clinical$`Organoids code`)]

colors$`MSI status` <- colors$msi_status
top_annotation <- HeatmapAnnotation(df = top_df,show_annotation_name = T,
                                        name = "MSI status",
                                        gap = unit(0.5, "mm"),
                                        border = FALSE,
                                        annotation_height = unit(3, "mm"),
                                        annotation_name_gp = gpar(fontsize=12),
                                        simple_anno_size_adjust = TRUE,
                                        annotation_name_side = "right",
                                        show_legend = FALSE,
                                        col = colors)


hm <-  Heatmap(matrix = heatdata,
               top_annotation = top_annotation,
               name = "z-score",
               clustering_distance_columns = "pearson",
               clustering_distance_rows = "pearson",
               heatmap_legend_param = list(title_position = "leftcenter-rot", at = c(-3,-2,-1,0,1,2,3),
                                           title_gp = gpar(fontsize = 18, fontface = "plain"), labels_gp = gpar(fontsize = 15)),
               show_column_dend = TRUE,
               show_row_dend = TRUE,
               column_names_max_height = unit(0.2, "npc"),
               row_names_max_width = unit(0.45, "npc"),
               show_heatmap_legend = TRUE,
               row_names_gp = gpar(fontsize = 12),
               column_names_gp = gpar(fontsize = 16),
               cluster_rows = T,
               cluster_columns = T)
```

Export as Figure 2C.
```{r, echo = FALSE}
pdf(file.path(params$paper_figures, "Figure2C.pdf"), width = 13, height = 13, onefile = TRUE)
draw(hm, heatmap_legend_side = "left")
dev.off()
```


### Protein complexes:

```{r}
#z-score across organoids:
prot.scaled <- scale(prot.avg)

# Load table from Ori et al., Genome Biology (2016)
file_protcompl <- "../../tables/protComplexes_13059_2016_912_MOESM2_ESM-1.xlsx"
prot.compl <- read_excel(file_protcompl, 1)

#select only human protein complexes:
prot.compl <- prot.compl[which(prot.compl$`Original source species` == "Human"), ]

#get list of ensembl IDs for complexes:
ensIDs <- as.list(strsplit(prot.compl$`Member identifiers (human Ensembl gene)`, '\\s+'))
names(ensIDs) <- prot.compl$`Complex ID`

#convert ensembl IDs to gene names:
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

prot_list <- vector(mode="list", length=length(ensIDs))
supp_table <- data.frame(`Complex ID` = character(),
                         HGNC = character(), 
                         nProt = integer(),
                         dProt = character(),
                         stringsAsFactors = FALSE)

for (i in 1:length(ensIDs)) {
  name <- names(ensIDs)[i]
  IDs <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","hgnc_symbol"), values=ensIDs[[i]], mart= mart)
  values <- prot.avg[rownames(prot.avg) %in% IDs$hgnc_symbol, ]
  
  if (! is.null(dim(values)[1]) && dim(values)[1] >= 5) {
    prot_list[[i]] <- values
    names(prot_list)[[i]] <- name
    hgnc <- paste(IDs$hgnc_symbol, collapse=" ")
    dProt <- paste(rownames(prot.avg)[rownames(prot.avg) %in% IDs$hgnc_symbol], collapse=" ")
    supp_table[i,] <- cbind(name, hgnc, dim(values)[1], dProt)
  }
  else {
    prot_list[[i]] <- NULL
    supp_table[i,] <-cbind(name, NA, NA, NA)
  }
}

supp_table <- merge(prot.compl, supp_table, by.x = "Complex ID", by.y ="Complex.ID")
supp_table <- supp_table[! is.na(supp_table$nProt),]

# delete empty entries:
prot_list <- prot_list[lapply(prot_list,length)>0]

# calculate correlation:
med_cor_org <- lapply(prot_list, cor, use="na.or.complete")
med_cor_org <- lapply(med_cor_org, function(x) {diag(x) <- NA; x})
med_cor_org <- lapply(med_cor_org, median)

med_cor_prot <- lapply(prot_list, t)
med_cor_prot <- lapply(med_cor_prot, cor, use="na.or.complete")
med_cor_prot <- lapply(med_cor_prot, function(x) {diag(x) <- NA; x})
med_cor_prot <- lapply(med_cor_prot, median, na.rm=TRUE)
```

```{r}
median.correlation <- as.data.frame(unlist(med_cor_prot))
median.correlation$`Complex ID` <- rownames(median.correlation)
names(median.correlation) <- c('median corr', 'Complex ID')
supp_table <- merge(supp_table, median.correlation, by="Complex ID")

write_tsv(supp_table, file=file.path(params$resultdir, "SuppTable1.txt"))
```

Visualize correlation in sorted barplot:
```{r,  fig.height=20, fig.width=15}
med_cor_prot <- as.data.frame(unlist(med_cor_prot))
colnames(med_cor_prot)[1] <- "med_cor_prot"

med_cor_prot$complex <- as.character(lapply(rownames(med_cor_prot), function(x) prot.compl$`Complex Name`[x == prot.compl$`Complex ID`]))
med_cor_prot$complex <- as.factor(med_cor_prot$complex)

p_med <- ggplot(med_cor_prot, aes(x= reorder(complex, med_cor_prot), y= med_cor_prot)) +
  geom_bar(stat="identity")+
  theme_minimal()+
  theme(axis.text.x = element_text(vjust = 0.5, hjust=1), axis.text.y = element_text(size=11), panel.grid.major = element_blank())+
  ylab("median Pearson's r")+
  xlab("")
p_med + coord_flip()
```

Select 26S Proteasome and plot members as heatmap:
```{r}
proteasome26S <- prot_list[[1]]
proteasome26S <- t(proteasome26S)
is.na(proteasome26S) <- !proteasome26S
```


Variances of members:
```{r,  fig.height=2, fig.width=10}
proteasome26S_variances <- apply(proteasome26S, 2, var, na.rm=TRUE)
proteasome26S_variances <- sort(proteasome26S_variances)
proteasome26S_variances <- as.data.frame(proteasome26S_variances)
proteasome26S_variances$subunit <- rownames(proteasome26S_variances)
names(proteasome26S_variances)[1] <- "variance"

proteasome26S_variances$subunit <- factor(proteasome26S_variances$subunit, levels=unique(as.character(proteasome26S_variances$subunit)) )

# add information about immuno- and constitutive proteasome to dataframe:
proteasome26S_variances$group <- rep('other sub-complex', dim(proteasome26S_variances)[1])
constitutiveproteasome = c('PSMB7', 'PSMB6', 'PSMB5')
immunoproteasome = c('PSMB8', 'PSMB9', 'PSMB10')
proteasome26S_variances$group[proteasome26S_variances$subunit %in% immunoproteasome] <- 'immunoproteasome'
proteasome26S_variances$group[proteasome26S_variances$subunit %in% constitutiveproteasome] <- 'constitutive proteasome'
proteasome26S_variances$group <- as.factor(proteasome26S_variances$group)

p <- ggplot(proteasome26S_variances, aes(x=subunit, y=variance, fill=group)) +
  geom_bar(stat="identity")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), text = element_text(size=30), panel.grid.major=element_blank(), panel.grid.minor=element_blank())+
  xlab("26S Proteasome subunits")

pdf(file.path(params$paper_figures, "Figure2F.pdf"), width = 20, height = 5, pointsize = 30)
p
dev.off()
```

Check protein expression level of PSMB8, PSMB9 and PSMB10 compared to all other proteins:
```{r,  fig.height=2.5, fig.width=8}
prot_16Ssubset <- prot.scaled[rownames(prot.scaled) %in% c('PSMB8', 'PSMB9', 'PSMB10'),]
hS1F <- Heatmap(prot_16Ssubset, cluster_rows = FALSE, name='z-score', show_column_dend = FALSE,
                heatmap_legend_param = list( labels_gp= gpar(fontsize=23), title_gp = gpar(fontsize=23)),
                column_names_gp = gpar(fontsize=30),
                row_names_gp = gpar(fontsize=30))
```

Export heatmap as SuppFigure 1F:
```{r}
pdf(file.path(params$paper_figures, "SuppFigure1F.pdf"), width = 20, height = 5, pointsize = 30)
draw(hS1F)
dev.off()
```