---
title: "01_mapDIA"
author: "Alexander Kirchmair"
params:
  data: ../../data
  resultsdir: ../../results
  SIGNOR: ../../tables/SIGNOR_all_data_23_04_21.tsv
  libdir: ../../lib
  tables: ../../tables
  figures: ../../results/figures
---

```{r setup, include=FALSE}

library(reshape2)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(paletteer)
library(openxlsx)
library(dplyr)
library(magrittr)

source(file.path(params$libdir, "phosphoproteomics_functions.R"))
source(file.path(params$libdir, "global_params.R"))

if (!dir.exists(params$resultsdir)){ dir.create(params$resultsdir) }
if (!dir.exists(file.path(params$resultsdir, "mapdia"))){ dir.create(file.path(params$resultsdir, "mapdia")) }
if (!dir.exists(params$figures)){ dir.create(params$figures) }

for (p in 1:length(params)){ do.call('Sys.setenv', params[p]) }

```


### 1. Preprocessing and mapDIA ----

```{r Preprocessing parameters}

phospho <- list()
phospho$thres$intensity <- 2000 # values below threshold are set to NA
phospho$thres$na <- 1 # max. allowed fraction of NAs per row
phospho$thres$nfrag <- 5 # minimum number of fragments
phospho$thres$max_na_ctrl <- 0 # max. allowed fraction of missing values in control group
phospho$thres$max_na_trt <- 0 # max. allowed fraction of missing values in treatment group
phospho$thres$min_trts_for_ctrl_imp <- Inf # number of treatment groups that must be measured to do imputation of the ctrl group
phospho$thres$max_na_keep <- 0 # maximum allowed fraction of missing values in a group

```


Data import
```{r Phosphoprotein data import}

phospho$raw <- read.delim2(file.path(params$data, "phosphoproteomics", "mapDIA_input_phospho.txt"), dec=".")
phospho$annotation <- phospho$raw[,colnames(phospho$raw) %in% c("ProteinName", "PeptideSequence", "FragmentIon", "RT")]
phospho$data <- phospho$raw[,!colnames(phospho$raw) %in% c("ProteinName", "PeptideSequence", "FragmentIon", "RT")]

phospho$design <- data.frame(row.names = colnames(phospho$data))
phospho$design$trt <- sapply(strsplit(rownames(phospho$design), split = "_"), function(tmp) tmp[1] )
phospho$design$trt[phospho$design$trt == "mTOR"] <- "mTORi"
phospho$design$treatment <- relevel(factor(phospho$design$trt), ref = "control")
phospho$design$rep <- sapply(strsplit(rownames(phospho$design), split = "_"), function(tmp) tmp[2] )
phospho$design$CRC <- NA
phospho$design$CRC[phospho$design$rep %in% 1:2] <- "CRC02"
phospho$design$CRC[phospho$design$rep %in% 3:4] <- "CRC02b"
phospho$design$CRC[phospho$design$rep %in% 5:6] <- "CRC03"
phospho$design$CRC[phospho$design$rep %in% 7:8] <- "CRC04"
phospho$design$CRC[phospho$design$rep %in% 9:10] <- "CRC13"
phospho$design$CRC[phospho$design$rep %in% 11:12] <- "CRC26"  
phospho$design$CRC[phospho$design$rep %in% 13:14] <- "CRC26LM"

phospho$design$organoid <- phospho$design$CRC
phospho$design$organoid[phospho$design$organoid == "CRC02b"] <- "CRC02"
phospho$design$crcname <- paste0(phospho$design$organoid, "_", phospho$design$treatment)
phospho$design$trtname <- paste0(phospho$design$treatment, "_", phospho$design$organoid)
phospho$design <- phospho$design[order(phospho$design$organoid),]
phospho$design <- phospho$design[order(phospho$design$treatment),]

phospho$design$name <- paste0(phospho$design$crcname, "_", phospho$design$rep)
phospho$data <- phospho$data[,rownames(phospho$design)]

```


Preprocessing
```{r Phosphoprotein preprocessing}

# Preprocess missing values
phospho$imputation <- impute_groups(data = phospho$data,
                                    design = phospho$design[colnames(phospho$data),],
                                    min_value = phospho$thres$intensity,
                                    max_na_ctrl = phospho$thres$max_na_ctrl ,
                                    max_na_trt = phospho$thres$max_na_trt,
                                    min_trts_for_ctrl_imp = phospho$thres$min_trts_for_ctrl_imp,
                                    max_na_keep = phospho$thres$max_na_keep)

phospho$imputed <- phospho$imputation$data

# Remove NA rows
phospho$row_na_fraction <- rowMeans(is.na(phospho$imputed))
phospho$annotation_na_filtered <- phospho$annotation[phospho$row_na_fraction  < phospho$thres$na,]
phospho$na_filtered <- phospho$imputed[phospho$row_na_fraction  < phospho$thres$na,]

# Remove peptides with too few fragments
phospho$annotation_na_filtered$ID <- paste0(phospho$annotation_na_filtered$PeptideSequence, "_", phospho$annotation_na_filtered$RT)
nfrag <- aggregate(. ~ ID, phospho$annotation_na_filtered[,c("ID", "FragmentIon")], FUN = length)
dimnames(nfrag) <- list(nfrag$ID, c("ID", "n"))
phospho$annotation_na_filtered$nfragments <- nfrag[phospho$annotation_na_filtered$ID,]$n
phospho$filtered <- phospho$na_filtered[phospho$annotation_na_filtered$nfragments >= phospho$thres$nfrag,]
phospho$annotation_filtered <- phospho$annotation_na_filtered[phospho$annotation_na_filtered$nfragments >= phospho$thres$nfrag,]

```


Preprocessing plots
```{r Phosphoprotein preprocessing plots, collapse = TRUE}

# Intensity filtering
ggdata <- melt(phospho$data, na.rm = TRUE)
ggdata$trt <- gsub(pattern = "_.*", "", ggdata$variable)
(gg1 <- ggplot(ggdata, aes(x = value, fill = trt)) +
  theme_classic(base_size = 20) + 
  geom_histogram(binwidth = 50, color = rgb(0,0,0)) +
  geom_vline(xintercept = phospho$thres$intensity, linetype="dashed") +
  scale_x_continuous(expand = c(0,NA), limits = c(0, phospho$thres$intensity*2.5)) +
  scale_y_continuous(expand = c(0,NA), limits = c(0, 10000)))
gg1

# Missing value filtering
(gg2 <- ggplot(data.frame("row_NA_fraction" = phospho$row_na_fraction), aes(x = row_NA_fraction)) +
  theme_classic(base_size = 20) +
  geom_histogram(binwidth = 0.05, color = "black") +
  geom_vline(xintercept = phospho$thres$na, linetype="dashed"))
gg2

# Fragment filtering
(gg3 <- ggplot(phospho$annotation_na_filtered, aes(x = nfragments)) +
  theme_classic(base_size = 20) +
  geom_histogram(binwidth = 1, color = "black")) +
  geom_vline(xintercept = phospho$thres$nfrag, linetype="dashed")
gg3

```


Write mapDIA input
```{r Phosphoprotein mapDIA input}

phospho$mapDIA_input_preprocessed <- data.frame(phospho$annotation_filtered, phospho$filtered, stringsAsFactors = FALSE)[,colnames(phospho$raw)]
write.table(phospho$mapDIA_input_preprocessed, file.path(params$resultsdir, "mapdia", "mapDIA_input_phospho.txt"),
            quote=FALSE, row.names=FALSE, sep="\t")

setmapDIAparams(design = phospho$design,
                formula = ~ CRC|treatment,
                params_file = file.path(params$resultsdir, "mapdia", "mapDIA_params_phospho.txt"),
                input_data = file.path(params$resultsdir, "mapdia", "mapDIA_input_phospho.txt"))
""
```


Run mapDIA
```{bash Phosphoprotein mapDIA, engine.opts = '-l', results = "hide"}

conda env create -f ../../../organoids/envs/mapDIA-env.yml -n organoids_mapdia
conda activate organoids_mapdia

cd $resultsdir/mapdia
mapDIA 'mapDIA_params_phospho.txt'

conda deactivate
conda remove --name organoids_mapdia --all -y

```



### 2. Postprocessing ----


Import all results from mapDIA
```{r Import results}

phosphodata <- list()
phosphoresults <- file.path(params$resultsdir, "mapdia")
phosphodata$log2data <- read.delim(file = file.path(phosphoresults, "log2_data.txt"), stringsAsFactors = FALSE) # fragment-level data
phosphodata$peptidelevels <- read.delim(file = file.path(phosphoresults, "peptide_level.txt"), stringsAsFactors = FALSE) # peptide-level data
phosphodata$proteinlevels <- read.delim(file = file.path(phosphoresults, "protein_level.txt"), stringsAsFactors = FALSE) # protein level data
phosphodata$DE.long <- read.delim(file = file.path(phosphoresults, "analysis_output.txt"), stringsAsFactors = FALSE) # DE output in long format

```



Postprocessing/annotation of phosphopeptide data
```{r Phosphopeptide annotation}

phosphodata$design <- phospho$design
phosphodata$imputation$impute <- data.frame(phospho$imputation$impute)[rownames(phospho$filtered),]
phosphodata$imputation$remove <- data.frame(phospho$imputation$remove)[rownames(phospho$filtered),]
phosphodata$imputation$annotation <- phospho$annotation_filtered

phosphodata$annotation <- data.frame(row.names = unique(phosphodata$log2data$Protein),
                                     protein = sapply(strsplit(unique(phosphodata$log2data$Protein), split = "__"), function(tmp) tmp[1] ),
                                     peptide = gsub("_", "", sapply(strsplit(unique(phosphodata$log2data$Protein), split = "__"), function(tmp) tmp[2] )), stringsAsFactors = FALSE)

phosphodata$annotation <- phosphodata$annotation[!grepl(pattern = "CON__", rownames(phosphodata$annotation)),]

```


```{r Phosphopeptide annotation}

geneslib <- read.delim(file.path(params$tables, "phosphoprot_to_gene_name.tsv"))
psiteres <- read.delim(file.path(params$tables, "phosphosites.tsv"))

rownames(geneslib) <- geneslib$uniprot
phosphodata$annotation$gene <- geneslib[phosphodata$annotation$protein,]$gene
phosphodata$annotation$synonyms <- geneslib[phosphodata$annotation$protein,]$synonyms

# For phosphopeptides with multiple matched proteins, split the proteins, get the phosphosite positions for each, and merge the rows back again
tmp.annotation <- phosphodata$annotation[,c("protein", "peptide", "gene")]
tmp.annotation$id <- 1:nrow(tmp.annotation)
tmp.annotation$protein.split <- tmp.annotation$protein
tmp.annotation.split <- c()
for (i in 1:nrow(tmp.annotation)){
  tmp <- tmp.annotation[i,,drop=FALSE]
  prot.all <- unique(unlist(strsplit(tmp$protein.split, split = ";")))
  for (j in 1:length(prot.all)){
    tmp.j <- tmp
    tmp.j$protein.split <- prot.all[j]
    tmp.annotation.split <- rbind(tmp.annotation.split, tmp.j)
  }
}

tmp.annotation.split$psite <- psiteres$psites
tmp.annotation.split$flankseqs <- psiteres$flankseqs

# get gene information
tmp.annotation.split$gene <- geneslib[tmp.annotation.split$protein.split,]$gene
tmp.annotation.split$gene[is.na(tmp.annotation.split$gene)] <- tmp.annotation.split$protein.split[is.na(tmp.annotation.split$gene)]
tmp.annotation.split$psite_name <- paste0(tmp.annotation.split$gene, "_", tmp.annotation.split$psite)
tmp.annotation.split$psite_name_protein <- paste0(tmp.annotation.split$protein.split, "_", tmp.annotation.split$psite)

# merge back by protein
tmp.annotation.split$psite <- sapply(tmp.annotation.split$id, function(tmpdup) paste(tmp.annotation.split$psite[ tmp.annotation.split$id == tmpdup ], collapse = ";") )
tmp.annotation.split$psite_name <- sapply(tmp.annotation.split$id, function(tmpdup) paste(tmp.annotation.split$psite_name[ tmp.annotation.split$id == tmpdup ], collapse = ";") )
tmp.annotation.split$psite_name_protein <- sapply(tmp.annotation.split$id, function(tmpdup) paste(tmp.annotation.split$psite_name_protein[ tmp.annotation.split$id == tmpdup ], collapse = ";") )

phosphodata$annotation$psite <- tmp.annotation.split[rownames(phosphodata$annotation),]$psite
phosphodata$annotation$psite_name <- tmp.annotation.split[rownames(phosphodata$annotation),]$psite_name
phosphodata$annotation$psite_name_protein <- tmp.annotation.split[rownames(phosphodata$annotation),]$psite_name_protein
phosphodata$annotation$flankseqs <- tmp.annotation.split[rownames(phosphodata$annotation),]$flankseqs

phosphodata$annotation$psite_name_pep <- phosphodata$annotation$psite_name
ix <- phosphodata$annotation$psite_name_pep %in% phosphodata$annotation$psite_name_pep[duplicated(phosphodata$annotation$psite_name_pep)] # all duplicates
i <- 1
phosphodata$annotation$psite_name_pep[ix] <- paste0(phosphodata$annotation$psite_name[ix], "_pep", i)
while (any(duplicated(phosphodata$annotation$psite_name_pep))){
    i <- i+1
    ix <- which(duplicated(phosphodata$annotation$psite_name_pep))
    phosphodata$annotation$psite_name_pep[ix] <- paste0(phosphodata$annotation$psite_name[ix], "_pep", i)
}

phosphodata$annotation$psite_name_protein_pep <- phosphodata$annotation$psite_name_protein
ix <- phosphodata$annotation$psite_name_protein_pep %in% phosphodata$annotation$psite_name_protein_pep[duplicated(phosphodata$annotation$psite_name_protein_pep)] # all duplicates
i <- 1
phosphodata$annotation$psite_name_protein_pep[ix] <- paste0(phosphodata$annotation$psite_name_protein[ix], "_pep", i)
while (any(duplicated(phosphodata$annotation$psite_name_protein_pep))){
    i <- i+1
    ix <- which(duplicated(phosphodata$annotation$psite_name_protein_pep))
    phosphodata$annotation$psite_name_protein_pep[ix] <- paste0(phosphodata$annotation$psite_name_protein[ix], "_pep", i)
}

phosphodata$pepdata <- phosphodata$peptidelevels[,!colnames(phosphodata$peptidelevels) %in% c("Protein", "Peptide", "nFragment")]
rownames(phosphodata$pepdata) <- phosphodata$peptidelevels$Protein
phosphodata$pepdata <- phosphodata$pepdata[!grepl(pattern = "CON__", rownames(phosphodata$pepdata)),]

# DE results annotation
phosphodata$DE.long$psite_name <- phosphodata$annotation[phosphodata$DE.long$Protein,]$psite_name
phosphodata$DE.long$labels <- phosphodata$annotation[phosphodata$DE.long$Protein,]$psite_name_pep

phosphodata$DE.long$sample <- sapply(strsplit(phosphodata$DE.long$Label2, split = "/", fixed = TRUE), function(tmp) tmp[1])
phosphodata$DE.long$trt <- sapply(strsplit(phosphodata$DE.long$sample, split = "_", fixed = TRUE), function(tmp) tmp[1])
phosphodata$DE.long$org <- sapply(strsplit(phosphodata$DE.long$sample, split = "_", fixed = TRUE), function(tmp) tmp[2])
phosphodata$DE.long$crc <- phosphodata$DE.long$org
phosphodata$DE.long$crc[phosphodata$DE.long$crc == "CRC02b"] <- "CRC02"
phosphodata$DE.long$group <- paste0(phosphodata$DE.long$crc, "_", phosphodata$DE.long$trt)

phosphodata$DE.pep <- phosphodata$DE.long[!grepl(pattern = "CON__", phosphodata$DE.long$Protein),]

ix_trtvsctrl <- sapply(gregexpr(pattern = "control", text = phosphodata$DE.pep$Label2), function(tmp) sum( tmp > 0 ) == 1)
phosphodata$DE.pep <- phosphodata$DE.pep[ix_trtvsctrl,]

# LFCs
phosphodata$lfc.pep <- dcast(phosphodata$DE.pep[,c("log2FC", "group", "labels")], formula = labels ~ group, value.var = "log2FC")
rownames(phosphodata$lfc.pep) <- phosphodata$lfc.pep$labels
phosphodata$lfc.pep$labels <- NULL

# FDRs
phosphodata$fdr.pep <- dcast(phosphodata$DE.pep[,c("FDR", "group", "labels")], formula = labels ~ group, value.var = "FDR")
rownames(phosphodata$fdr.pep) <- phosphodata$fdr.pep$labels
phosphodata$fdr.pep$labels <- NULL

phosphodata$DEdesign <- phosphodata$design[match(colnames(phosphodata$lfc), phosphodata$design$crcname),]
rownames(phosphodata$DEdesign) <- phosphodata$DEdesign$crcname
phosphodata$DEdesign$rep <- NULL

# Fraction of peptides with imputed values per phosphopeptide
impdf <- melt(aggregate(. ~ protein, data.frame("protein" = phosphodata$imputation$annotation$ProteinName, phosphodata$imputation$impute), FUN = mean))
impdf$id <- paste0(impdf$protein, "_", phosphodata$design[impdf$variable,]$treatment, "_", phosphodata$design[impdf$variable,]$CRC)
impdf <- aggregate(. ~ id, subset(impdf, select = c(id, value)), FUN = mean)
rownames(impdf) <- impdf$id
phosphodata$DE.pep$class2 <- paste0(phosphodata$DE.pep$Protein, "_", gsub("/.*", "", phosphodata$DE.pep$Label2))
phosphodata$DE.pep$class1 <- paste0(phosphodata$DE.pep$Protein, "_", gsub(".*/", "", phosphodata$DE.pep$Label2))
phosphodata$DE.pep$imputed_class2 <- impdf[phosphodata$DE.pep$class2,]$value
phosphodata$DE.pep$imputed_class1 <- impdf[phosphodata$DE.pep$class1,]$value

```


Selection of phosphopeptides containing the same phosphosite
```{r Phosphopeptide selection}

# Select multi-site peptides based on higher mean expression
rownames(phosphodata$peptidelevels) <- phosphodata$peptidelevels$Protein
tmpdata <- phosphodata$peptidelevels[rownames(phosphodata$annotation),rownames(phosphodata$design)]
phosphodata$annotation$mean <- rowMeans(tmpdata[,phosphodata$design$treatment == "control"])
phosphodata$annotation$select <- selectMultiples(v = phosphodata$annotation$mean, vnames = phosphodata$annotation$psite_name, FUN = max)

# Prepare results with phosphosite positions as unique identifiers
ids <- intersect(rownames(phosphodata$annotation), rownames(phosphodata$pepdata))
phosphodata$data <- phosphodata$pepdata[ids[phosphodata$annotation[ids,]$select],]
rownames(phosphodata$data) <- phosphodata$annotation[rownames(phosphodata$data),]$psite_name
phosphodata$DE <- phosphodata$DE.pep[phosphodata$DE.pep$labels %in% phosphodata$annotation$psite_name_pep[phosphodata$annotation$select],]
phosphodata$DE$labels <- phosphodata$DE$psite_name
phosphodata$DE$Peptide <- phosphodata$DE$Protein
phosphodata$DE$protein <- phosphodata$annotation[phosphodata$DE$Peptide,]$protein
phosphodata$DE$gene <- phosphodata$annotation[phosphodata$DE$Peptide,]$gene
phosphodata$lfc <- dcast(phosphodata$DE[,c("log2FC", "group", "labels")], formula = labels ~ group, value.var = "log2FC")
rownames(phosphodata$lfc) <- phosphodata$lfc$labels
phosphodata$lfc$labels <- NULL
phosphodata$fdr <- dcast(phosphodata$DE[,c("FDR", "group", "labels")], formula = labels ~ group, value.var = "FDR")
rownames(phosphodata$fdr) <- phosphodata$fdr$labels
phosphodata$fdr$labels <- NULL


# Prepare results with uniprot-id phosphosite positions as unique identifiers
ids <- intersect(rownames(phosphodata$annotation), rownames(phosphodata$pepdata))
phosphodata$protdata <- phosphodata$pepdata[ids[phosphodata$annotation[ids,]$select],]
rownames(phosphodata$protdata) <- phosphodata$annotation[rownames(phosphodata$protdata),]$psite_name_protein
phosphodata$protDE <- phosphodata$DE.pep[phosphodata$DE.pep$labels %in% phosphodata$annotation$psite_name[phosphodata$annotation$select],]
phosphodata$protDE$psite_name_protein <- phosphodata$annotation[phosphodata$protDE$Protein,]$psite_name_protein
phosphodata$protDE$labels <- phosphodata$protDE$psite_name_protein
phosphodata$protDE$Peptide <- phosphodata$protDE$Protein
phosphodata$protDE$protein <- phosphodata$annotation[phosphodata$protDE$Peptide,]$protein
phosphodata$protDE$gene <- phosphodata$annotation[phosphodata$protDE$Peptide,]$gene
phosphodata$prot.lfc <- dcast(phosphodata$protDE[,c("log2FC", "group", "labels")], formula = labels ~ group, value.var = "log2FC")
rownames(phosphodata$prot.lfc) <- phosphodata$prot.lfc$labels
phosphodata$prot.lfc$labels <- NULL
phosphodata$prot.fdr <- dcast(phosphodata$protDE[,c("FDR", "group", "labels")], formula = labels ~ group, value.var = "FDR")
rownames(phosphodata$prot.fdr) <- phosphodata$prot.fdr$labels
phosphodata$prot.fdr$labels <- NULL

# Prepare results with flanking sequences as unique identifiers
phosphodata$annotation$flankseqs[is.na(phosphodata$annotation$flankseqs)] <- "NA"
phosphodata$annotation$flank.select <- selectMultiples(v = phosphodata$annotation$mean, vnames = phosphodata$annotation$flankseqs, FUN = max)
phosphodata$flank.DE <- phosphodata$DE.pep[phosphodata$DE.pep$labels %in% phosphodata$annotation$psite_name_pep[phosphodata$annotation$flank.select],]
phosphodata$flank.DE$flankseqs <- phosphodata$annotation[phosphodata$flank.DE$Protein,]$flankseqs
phosphodata$flank.DE$labels <- phosphodata$flank.DE$flankseqs
phosphodata$flank.DE$Peptide <- phosphodata$flank.DE$Protein
phosphodata$flank.DE$protein <- phosphodata$annotation[phosphodata$flank.DE$Peptide,]$protein
phosphodata$flank.DE$gene <- phosphodata$annotation[phosphodata$flank.DE$Peptide,]$gene
phosphodata$flank.lfc <- dcast(phosphodata$flank.DE[,c("log2FC", "group", "labels")], formula = labels ~ group, value.var = "log2FC")
rownames(phosphodata$flank.lfc) <- phosphodata$flank.lfc$labels
phosphodata$flank.lfc$labels <- NULL
phosphodata$flank.fdr <- dcast(phosphodata$flank.DE[,c("FDR", "group", "labels")], formula = labels ~ group, value.var = "FDR")
rownames(phosphodata$flank.fdr) <- phosphodata$flank.fdr$labels
phosphodata$flank.fdr$labels <- NULL

```


LFC comparison
```{r}

lfcs <- unique(phosphodata$design$CRC) %L>% function(crc){
  grep("control", unique(subset(phosphodata$design, CRC == crc)$trtname), value = TRUE, invert = TRUE) %S>% function(trt){
    tmpsamples <- rownames(phosphodata$design)[phosphodata$design$trtname == trt]
    tmpctrl <- rownames(phosphodata$design)[phosphodata$design$treatment == "control" & phosphodata$design$CRC == crc]
    log2( rowMeans(phosphodata$peptidelevels[,tmpsamples, drop = FALSE]) / rowMeans(phosphodata$peptidelevels[,tmpctrl, drop = FALSE]))
  }
}
lfcs <- lfcs %>% Reduce(f = cbind)
rownames(lfcs) <- phosphodata$annotation[rownames(lfcs),"psite_name"]

df <- melt(lfcs)
df$sample <- factor(df$Var2, levels = sort(unique(df$Var2)))
df$ids <- paste0(df$Var1, "_", df$Var2)
tmpids <- paste0(phosphodata$DE$psite_name, "_", phosphodata$DE$trt, "_", phosphodata$DE$crc)
df$mapdia <- phosphodata$DE$log2FC[match(df$ids, tmpids)]
df$FDR_mapdia <- phosphodata$DE$FDR[match(df$ids, tmpids)]

gg <- ggplot(df, aes(x = value, y = mapdia, color = -log10(FDR_mapdia))) +
  theme_bw() +
  geom_point(size = 0.05) + 
  scale_color_gradient(low = rgb(0.2,0.2,0.2,0.2), high = rgb(0.9,0,0), na.value = rgb(0,0,0,0.1)) +
  facet_wrap(~ sample, scales = "fixed", ncol = 6) + xlab("lfc") + ylab("LFC_mapdia")

 gg

```



Volcanoplot
```{r Volcanoplot}

phosphodata$DE$gglabels <- phosphodata$DE$labels
phosphodata$DE$descore <- abs(as.numeric(scale(-log10(phosphodata$DE$FDR+1))) * as.numeric(scale(phosphodata$DE$log2FC)))
phosphodata$DE$gglabels[phosphodata$DE$descore < sort(phosphodata$DE$descore, decreasing = TRUE)[50]] <- ""

gg <- ggplot(phosphodata$DE, aes(x=log2FC, y=-log10(FDR), color = Label2, label = gglabels)) +
  theme_bw(base_size = 26) +
  geom_point(alpha=0.8, shape = 19, size = 1, stroke=0) +
  geom_text_repel() +
  scale_color_paletteer_d("ggsci::default_igv") +
  guides(color = FALSE) +
  ggtitle(label = NULL, subtitle = paste0("FDR < 0.05: ", sum(phosphodata$DE$FDR <= 0.05), " phosphosites"))
phosphodata$DE$gglabels <- NULL

gg

```

PCA plot
```{r PCA}

anno_df <- phosphodata$design[,c("treatment", "organoid")]
rownames(anno_df) <- phosphodata$design$name
anno_df$treatment <- factor(anno_df$treatment)
anno_df$treatment <- relevel(anno_df$treatment, ref = "control")

pcadata <- t(phosphodata$data)
rownames(pcadata) <- phosphodata$design[rownames(pcadata),]$name
pca <- prcomp(pcadata[,apply(pcadata, 2, var, na.rm = TRUE) != 0], center = T, scale. = T)
pca_df <- data.frame(pca$x[rownames(anno_df),], anno_df)
variance_explained <- summary(pca)$importance["Proportion of Variance", ]

ggpca <- ggplot(pca_df, aes(x=PC1, y=PC2, color = organoid, shape = treatment)) + 
  geom_point(size = 4, stroke = 1.5) + 
  theme_bw(base_size = 18) + 
  theme(panel.border = element_rect(colour = "black", fill = "transparent", size = 1)) +
  scale_color_manual(values = colors$organoid[unique(pca_df$organoid)]) + 
  scale_shape_manual(values = c(20, 0:c(length(unique(pca_df$treatment))-2))) +
  xlim(min(pca_df$PC1), 1.1*max(pca_df$PC1)) + 
  xlab(paste0("PC1 (", 100*round(variance_explained[1], 2), "%)")) + 
  ylab(paste0("PC2 (", 100*round(variance_explained[2], 2), "%)")) + 
  labs(color = "organoid", shape = "treatment") +
  coord_fixed(ratio = 1)

ggsave(plot = ggpca, device = "pdf",
       filename = file.path(params$figures, "Fig3a.pdf"),
       units = "px", width = 2500, height = 2000, dpi = 300,bg = "transparent")

```


LFC heatmap of signif. phosphosites
```{r LFC heatmap}

heatdata <- phosphodata$lfc[phosphodata$fdr %row<% 0.05,]
heatdata[is.na(heatdata)] <- 0

top_df <- phosphodata$design[match(colnames(heatdata), phosphodata$design$crcname), c("treatment", "organoid", "crcname")]
rownames(top_df) <- top_df$crcname
top_df$crcname <- NULL
colnames(top_df) <- c("treatment", "organoid")

heatmap(heatdata,
        rownames_width = 0.4,
        scale_rows = FALSE,
        top_df = top_df,
        plot_top_legend = TRUE,
        top_df_colors = colors,
        top_df_height = 3,
        annofontsize = 18,
        coltitle = "",
        rowlabsize = 0,
        cluster_rows = TRUE,
        plot_row_clusters = TRUE,
        coldend_size = 0.06,
        rowdend_size = 0.06,
        colorscale_legend_title = "lfc",
        anno_leg_size = 1.5,
        legend_separate_file = F,
        col_leg_height = 0.7,
        col_leg_width = 1,
        leg_fontsize = 18, 
        colorscale = colorRamp2(c(-3,0,3), c("blue", "white", "red")),
        bg = rgb(1,1,1), width = 5000, height = 4000, 
        scale_margins = c(1,1.5,1,1),
        use_raster = T,
        pdffile = file.path(params$figures, "SuppFig2c"))

```



Target site plots 
```{r Treatment target phosphosites}

signor <- read.csv(file.path(params$SIGNOR), sep = "\t", stringsAsFactors = FALSE)
signor %<>% subset(TAX_ID %in% c(9606, -1)) %>% signor2net() %>%  subset(PSITE != "")
target_sites <- trt_targets %L>% function(tmp){ sort(unique(signor$PSITE_NAME[signor$from %in% tmp])) }

# add known sites
target_sites$TNFalpha <- base::union(target_sites$TNFalpha, "MAPK14_Y182")
target_sites$TBKi <- base::union(target_sites$TBKi, "MAP3K1_S923")
target_sites$mTORi <- base::union(target_sites$mTORi, "IRS2_T527")
target_sites$PI3Ki <- base::union(target_sites$PI3Ki, "IRS2_T527")


measured_sites <- names(target_sites) %L>% function(tmp){intersect(subset(phosphodata$DE, trt == tmp)$psite_name, target_sites[[tmp]])}
measured_sites <- measured_sites[measured_sites %S>% length() > 0]

phosphodata$DE$crc <- factor(phosphodata$DE$crc)
phosphodata$DE$DE <- factor(ifelse(phosphodata$DE$FDR <= 0.05, "sig", "not sig"))

site_plots <- names(measured_sites) %L>% function(tmp){
  
  plotdata <- subset(phosphodata$DE, psite_name %in% measured_sites[[tmp]] & trt == tmp)
  colnames(plotdata)[colnames(plotdata) == "labels"] <- "phosphosite"
  
  ggplot(data = plotdata, aes(x = phosphosite, y = log2FC, color = log2FC, fill = log2FC, shape = DE)) +
    theme_bw(base_size = 18) + xlab(NULL) +
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
          legend.title = element_blank(),
          legend.margin = margin(0.5, 1, 1, 1),
          strip.background =  element_rect(colour = NA, fill = rgb(0.85,0.85,0.85), size=1),
          axis.text.x = element_text(color = rgb(0,0,0), size = 7, angle = 90, hjust = 1, vjust = 0.5,
                                     margin = margin(t = 0, r = 0, b = 0.02, l = 0, unit = "npc")),
          plot.title = element_text(hjust = 0.5),
          plot.margin = margin(5, 5, 5, 5, unit = "pt"),
          legend.position = "right") +
    geom_segment(aes(xend=phosphosite, y=0, yend=log2FC), size = 1.3) +
    geom_hline(yintercept=0, linetype="dashed", color = rgb(0,0,0)) + 
    geom_point(size = 4, colour="black") +
    scale_shape_manual(values = c("sig" = 23, "not sig" = 21), drop = FALSE) +
    ggtitle(tmp) + 
    facet_wrap(. ~ crc, nrow = 1, scales = "free_x", drop = FALSE) + 
    scale_color_gradient2(low = "#5106b2", mid = rgb(0.94,0.94,0.94), high = "#f2c80e", midpoint = 0) +
    scale_fill_gradient2(low = "#5106b2", mid = rgb(0.94,0.94,0.94), high = "#f2c80e", midpoint = 0) +
    ylim(1.2*c(floor(min(plotdata$log2FC)), max(ceiling(plotdata$log2FC), 0.2))) +
    guides(color = "none", fill = "none")
}

site_plot <- plot_grid(plotlist = site_plots, ncol = 1, align = "hv")
ggsave(path = params$figures, filename = "SuppFig2b.pdf",
       plot = site_plot, device = "pdf", units = "in", width = 17, height = 18, dpi = 300)

```


```{r Export data}

if (!file.exists(file.path(params$data, "phosphoproteomics", "sample_info.tsv"))){
  write.table(x = data.frame(sample = rownames(phosphodata$design), select(phosphodata$design, -c(trt, CRC, crcname, trtname)))[colnames(phosphodata$log2data)[-(1:3)],],
              file = file.path(params$data, "phosphoproteomics", "sample_info.tsv"),
              quote = FALSE, row.names = FALSE, sep = "\t")
}

if (!file.exists(file.path(params$data, "phosphoproteomics", "mapDIA_results_phospho.tsv"))){
  write.table(x = select(phosphodata$DE, -c(labels, sample, org, group, class2, class1, imputed_class2, imputed_class1, descore, DE)),
              file = file.path(params$data, "phosphoproteomics", "mapDIA_results_phospho.tsv"),
              quote = FALSE, row.names = FALSE, sep = "\t")
}

```


```{r}

saveRDS(phosphodata, file.path(params$resultsdir, "01_phosphodata.rds"))

```



